\acknowledgments{% \vskip2pc
  % {\narrower\noindent
\begin{singlespace}

I would like to express immense gratitude to Mary Lidstrom, who taught me so much about biology and metabolic engineering, then supported my transition to computational biology and data science.
Mary was an incredible mentor for every phase of my PhD work, providing brilliant ideas on unfamiliar new data both before and after my pivot to computational science.
Every interaction was positive, every meeting productive, and she was incredibly accessible while simultaneously serving as the Vice Provost of Research for the University of Washington.
It was always clear that my and other lab members' best interests for scientific and professional development were her top priority.
Mary will forever remain one of my top role models for science and leadership.

I also  would like to thank David Beck for his computational mentorship throughout my PhD, and particularly for the last two years when we worked so closely together.
His incredible breadth and depth of knowledge, and his eagerness to share it with students are unparalleled.
It was Dave's enthusiasm for data science that led me to the Advanced Data Science Option, which has already had positive impacts on my career.

I am also thankful for the dozens of people who lent support and collaborated on ambitious wet lab experiments, most of which are not described in this thesis.
I thank the Formolase project team, particularly Amanda Smith, David Baker, Adam Wargacki, and Justin Siegel.
I thank the numerous Lidstrom Lab members for support around microbiology, molecular biology, and omics techniques.
I thank Marina Kalyuzhnaya for initial mentorship when I joined the lab, and Ludmila Chistoserdova for regular advice concerning methylotrophy.

I thank my family for guiding me toward engineering at a young age, and the support through these 25 years of education.

I also thank the community around the Data Science Studio and Advanced Data Science Option at UW.
%The impact of these programs on my training and career trajectory is substantial.

Finally, I thank my committee members for support throughout a particularly diverse PhD experience.

\end{singlespace}
% Internship companies.
}

