\chapter*{Glossary} % \chapter*{Glossary} <-- the star turns off the chapter numbering and removes it from the index.
\renewcommand*{\arraystretch}{1.4} % space between entries % http://tex.stackexchange.com/questions/135593/adding-row-spacing-to-a-longtable

\begin{singlespace}
\begin{longtable}{ p{.20\textwidth}  p{.80\textwidth} }

\textbf{AMI} & Amazon machine image, a master image for the creation of virtual servers (known as EC2 instances)
                in the Amazon Web Services (AWS) environment. \\
\textbf{ANI} & Average nucleotide idenity, a measure of similarity between organisms. \\
\textbf{assembly} &  \\
\textbf{AWS} & Amazon Web Services, the leading cloud computing platform \\
\textbf{bin} &  \\
\textbf{binning} &   \\ % "clustering"?
\textbf{BWA} & Burrows-Wheeler Aligner, a tool for maping reads to reference DNA sequences \\
\textbf{CCA} & Cannonical correlation analysis \\
\textbf{CheckM} &  \\
\textbf{contig} & A contiguous DNA sequence.  In this case, they are the longer DNA stretches identified from assembling Illumina reads. \\
\textbf{C1} & Single-carbon compounds such as methanol, \ce{CO_2}, or formate \\
\textbf{coverage} & number of reads that include a given nucleotide in the reconstructed sequence \\
\textbf{cross-validation} &  \\
\textbf{d} &  \\
\textbf{EC2} & a virtual server in Amazon’s Elastic Compute Cloud (EC2) \\
\textbf{EFS} &  \\
\textbf{Elviz} &  \\
\textbf{FASTA} &  TODO: note that blocks of sequence can be the DNA of a single contig, a single gene, or the nucleotide of a single gene.  \\
\textbf{fastq} &  \\
\textbf{feature} &  \\
\textbf{GC} & DNA bases.  When used in terms like "high GC", it indicates high fraction of G, C nucleotides over A, T nucleotides. \\
\textbf{genomics} & the collective characterization and quantification of pools of DNA molecules \\
\textbf{genome bin} & a collection of contigs used to approximate a single organism's genome.  Often referred to as "bin". \\
\textbf{graph} &  \\
\textbf{held-out set} &  \\
\textbf{hyperparameter} &  \\
\textbf{i.i.d.} & Independent and identically distributed.
    Each random variable has the same probability distribution as the others and all are mutually independent \\
\textbf{isolate} & \\
\textbf{JGI} & the Joint Genome Institute, a U.S. Department of Energy laboratory which provided sequencing
                services for the metagenomics work of this thesis \\
\textbf{kb} & kilobase. A unit of DNA (or RNA) length, corresponding to 1000 base pairs. \\
\textbf{mapping} & Mapping of reads means XXX \\
\textbf{machine learning} &  \\
\textbf{metabat} &  \\
\textbf{metagenome} &  \\
\textbf{metatranscriptome} &  \\
\textbf{methanotroph} &  \\
\textbf{methylotroph} &  \\
\textbf{MyCC} &  \\
\textbf{N} &  \\
\textbf{N50} & a weighted median statistic for assemblies.  %http://biology.stackexchange.com/questions/34122/an-example-for-n50-why-do-we-need-it
    Defined such that 50\% of the entire assembly is contained in contigs or scaffolds equal to or larger than this value. \\
    %the length for which the collection of all contigs of that length, or longer, contains at least half of the total of the
    %            lengths of the contigs in the assembly” \\ % https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4426941/pdf/bbi-9-2015-075.pdf (cites 49, 50)
\textbf{OB3b} & A RuMP cycle methanotroph, that is the subject of study in Chapter \ref{chapter:A}  \\
\textbf{omics} & refers to fields of study ending in \textit{-omics}, such as genomics and transcriptomics, which quantify pools of biological molecules \\
\textbf{partial correlation} &  \\
\textbf{PCA} principle components analysis &  \\
\textbf{PhyloPhlAn} &  \\
\textbf{Prokka} &  \\
\textbf{read} & a single short DNA sequence of a larger high-throughput sequencing data set. \\
\textbf{regularization} &  \\
\textbf{taxa} &  \\
\textbf{taxonomy} &  \\
\textbf{tetranucleotide frequency} & The frequency of 4-nucleotide sequences in DNA, which is conserved for an organism. \\
\textbf{training set} &  \\
\textbf{transcriptomics} & the collective characterization and quantification of pools of RNA molecules (transcripts) \\
\textbf{16S} &  \\

\end{longtable}
\end{singlespace}
